---
title: "Compare package versions on Bioconductor and GitHub"
author: "Charlotte Soneson"
date: today
format: 
    html:
        embed-resources: true
        toc: true
        theme: united
---

The goal of this analysis is to compare the package version of Bioconductor 
packages between the Bioconductor git server and the user repository. At the 
moment, for simplicity we limit ourselves to packages that have a GitHub 
repository indicated in the `URL` field in their DESCRIPTION file (a bit less
than half of the packages in the current Bioconductor devel branch). 

# Load packages

```{r}
suppressPackageStartupMessages({
    library(BiocPkgTools)
    library(DT)
    library(dplyr)
})
```

# Retrieve data

```{r, warning=FALSE}
# path to output file
outfile <- "pl_with_github_url.rds"

# set to TRUE to force retrieval of data even if outfile exists
force_retrieval <- FALSE

# helper functions
# ... get the GitHub repository from the URL field (which can contain multiple links)
.get_github_repo <- function(url) {
    # split string by spaces, commas, newlines
    url <- trimws(unlist(strsplit(url, ", |\n|,\n|,")))
    # grab only the entry pointing to a GitHub repository
    url <- url[grepl("github.com", url)]
    # extract repository name (if there are multiple GitHub repositories, 
    # choose the first one)
    sub("/$", "", sub("^.*github.com/(.+)$", "\\1", url)[1])
}

# ... attempt to read a DESCRIPTION file from the devel, main or master 
#     branch of a repo, and extract the version. If no DESCRIPTION file is 
#     found in these branches, return an empty string
.get_github_version <- function(repo) {
    desc <- tryCatch({
        read.dcf(url(paste0("https://raw.githubusercontent.com/", repo,
                            "/refs/heads/devel/DESCRIPTION")))
    }, error = function(e) {
        NULL
    })
    if (is.null(desc)) {
        desc <- tryCatch({
            read.dcf(url(paste0("https://raw.githubusercontent.com/", repo,
                                "/refs/heads/main/DESCRIPTION")))
        }, error = function(e) {
            NULL
        })
    }
    if (is.null(desc)) {
        desc <- tryCatch({
            read.dcf(url(paste0("https://raw.githubusercontent.com/", repo,
                                "/refs/heads/master/DESCRIPTION")))
        }, error = function(e) {
            NULL
        })
    }
    if (is.null(desc)) {
        version <- ""
    } else {
        version <- desc[1, "Version"]
    }
    version
}

# retrieve information about all Bioc packages, and check how many have a 
# GitHub repository in the URL field (we only analyze these further)
pl <- BiocPkgTools::biocPkgList()
table(grepl("github.com", pl$URL))
    
if (!file.exists(outfile) || force_retrieval) {
    # retrieve version for packages with GitHub URL
    pl_with_github_url <- pl |> 
        dplyr::filter(grepl("github.com", URL)) |>
        dplyr::mutate(Repo = vapply(URL, .get_github_repo, "")) |>
        dplyr::mutate(GitHubVersion = vapply(Repo, .get_github_version, "")) |>
        dplyr::select(Package, Repo, Version, GitHubVersion)
    
    saveRDS(pl_with_github_url, outfile)
} else {
    pl_with_github_url <- readRDS(outfile)
}

# check for NA values or wrongly processed repositories
# repositories where the GitHub version is not found could be e.g. those where
# the package lives in a subfolder of the repository
table(is.na(pl_with_github_url$Repo))
unname(grepv("github", pl_with_github_url$Repo))
table(pl_with_github_url$GitHubVersion == "")

DT::datatable(pl_with_github_url, 
              filter = list(position = "top", clear = FALSE),
              options = list(scrollX = TRUE, pageLength = 20))

# only keep packages where the GitHub version could be retrieved
pl_with_github_url <- pl_with_github_url |>
    dplyr::filter(GitHubVersion != "")
dim(pl_with_github_url)

# convert the version numbers to numeric versions
pl_with_github_url$GitHubVersion <- 
    package_version(pl_with_github_url$GitHubVersion)
pl_with_github_url$Version <- 
    package_version(pl_with_github_url$Version)
```

# Compare versions

```{r}
# packages with matching version
table(pl_with_github_url$Version == 
          pl_with_github_url$GitHubVersion)

# packages with newer version on Bioconductor
table(pl_with_github_url$Version >
          pl_with_github_url$GitHubVersion)

# packages with newer version on GitHub
table(pl_with_github_url$Version <
          pl_with_github_url$GitHubVersion)
DT::datatable(pl_with_github_url |>
                  dplyr::filter(Version < GitHubVersion) |>
                  dplyr::mutate(Version = as.character(Version),
                                GitHubVersion = as.character(GitHubVersion)) |>
                  dplyr::select(-Repo),
              filter = list(position = "top", clear = FALSE),
              options = list(scrollX = TRUE, pageLength = 20))
```

# Focus on packages with newer version on Bioconductor

```{r}
pl_newer_on_bioc <- pl_with_github_url |> 
    dplyr::filter(Version > GitHubVersion) |>
    dplyr::mutate(
        major_diff = as.numeric(Version[, 1]) - as.numeric(GitHubVersion[, 1]), 
        minor_diff = as.numeric(Version[, 2]) - as.numeric(GitHubVersion[, 2]),
        patch_diff = as.numeric(Version[, 3]) - as.numeric(GitHubVersion[, 3]),
        github_major_version_0 = GitHubVersion[, 1] == "0"
    )
dim(pl_newer_on_bioc)

DT::datatable(pl_newer_on_bioc |>
                  dplyr::mutate(Version = as.character(Version),
                                GitHubVersion = as.character(GitHubVersion)) |>
                  dplyr::select(-Repo),
              filter = list(position = "top", clear = FALSE),
              options = list(scrollX = TRUE, pageLength = 20))

# check whether packages with newer version on Bioconductor have patch 
# version 0 (this could, but doesn't have to, indicate that no changes have 
# been made to the package code upstream - at least not in the current branch)
table(as.character(pl_newer_on_bioc$Version[, 3]))

# check the difference in the major/minor version
# stratify based on whether the package version on GitHub is corresponding 
# to a "new submission" (major version = 0) or not
table(major_diff = pl_newer_on_bioc$major_diff, 
      minor_diff_sign = sign(pl_newer_on_bioc$minor_diff), 
      github_major_version_0 = pl_newer_on_bioc$github_major_version_0)

# some unexpected cases
pl_newer_on_bioc |>
    dplyr::filter(github_major_version_0 & 
                      sign(minor_diff) != -1) |>
    dplyr::select(Package, Version, GitHubVersion) |>
    dplyr::mutate(Version = as.character(Version),
                  GitHubVersion = as.character(GitHubVersion))
```

# Conclusions

* Many package maintainers appear not to sync their GitHub repositories to pull down changes made upstream (e.g. version bumps or other edits made by the core team). However, it is important to note that this doesn't mean that the package content is different - a more thorough analysis would be needed to check if this is actually the case. Among the packages with a newer version available on Bioconductor, the majority have patch version equal to 0 on Bioconductor, indicating that no changes were made to the package in the current (devel) branch. Of course, this does not mean that no changes were made upstream since the last synchronization with the local GitHub repository.  
* Most local changes are pushed upstream - only few packages have a higher version number on GitHub, and in most cases the difference is small (likely indicating ongoing development).

# Session info

```{r}
sessionInfo()
```

